---
title: "Continual Reassessment Method: Simulation Practice"
author: ""
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 5,
  fig.align = "center"
)
source("R/crm_functions.R")
```

# Purpose

This coding exercise implements a one-parameter Bayesian continual reassessment
method (CRM). It covers data generation, multiple skeletons, grid and MCMC
posterior computation, sequential dose assignment, BMA-CRM, and operating
characteristics under several true dose-toxicity scenarios.

The code is intentionally transparent and uses base R. It is for learning and
is **not validated software for conducting a clinical trial**.

# Design settings and data-generating scenarios

We use six ordered dose levels, a target DLT probability of 0.25, cohorts of
three, and 24 patients. The true MTD is the dose whose DLT probability is
closest to the target.

```{r design-settings}
target <- 0.25
n_dose <- 6L
n_patients <- 24L
cohort_size <- 3L
prior_mtd <- 3L

scenarios <- list(
  "MTD at dose 2" = c(0.08, 0.23, 0.38, 0.55, 0.68, 0.80),
  "MTD at dose 4" = c(0.04, 0.08, 0.14, 0.24, 0.38, 0.55),
  "MTD at dose 5" = c(0.02, 0.05, 0.09, 0.14, 0.23, 0.36)
)

scenario_table <- data.frame(
  scenario = rep(names(scenarios), each = n_dose),
  dose = rep(seq_len(n_dose), times = length(scenarios)),
  true_dlt_probability = unlist(scenarios, use.names = FALSE)
)
knitr::kable(scenario_table, digits = 2)
```

# CRM working model

For skeleton probabilities \(p_1 < \cdots < p_J\), we use the power model

\[
\Pr(Y=1\mid d_j,\alpha)=p_j^{\exp(\alpha)},
\qquad \alpha\sim N(0,2),
\]

where 2 is the prior variance. After each cohort, the next dose is the level
whose posterior mean DLT probability is closest to the target. Dose movement is
limited to one level per cohort.

Sequential decisions use deterministic grid integration. MCMC is demonstrated
separately, avoiding simulation decisions that depend on Monte Carlo noise.

# Lee-Cheung indifference-interval skeletons

Lee and Cheung (2009) specify an indifference interval
\((\phi-\delta,\phi+\delta)\) around target \(\phi\). Given prior MTD location
\(\nu\), set \(p_\nu=\phi\) and use

\[
p_{j-1}=\exp\left\{
\frac{\log(\phi-\delta)\log(p_j)}{\log(\phi+\delta)}
\right\},
\]

\[
p_{j+1}=\exp\left\{
\frac{\log(\phi+\delta)\log(p_j)}{\log(\phi-\delta)}
\right\}.
\]

A smaller \(\delta\) gives a more closely spaced skeleton. We compare three
prespecified values and assess them through finite-sample simulations.

```{r construct-skeletons}
deltas <- c(0.04, 0.07, 0.10)
skeletons <- lapply(deltas, function(delta) {
  lee_cheung_skeleton(target, delta, n_dose, prior_mtd)
})
names(skeletons) <- paste0("CRM delta=", sprintf("%.02f", deltas))

skeleton_table <- data.frame(
  dose = seq_len(n_dose),
  do.call(cbind, skeletons),
  check.names = FALSE
)
knitr::kable(skeleton_table, digits = 3)
```

```{r skeleton-plot, fig.cap="Three Lee-Cheung skeletons."}
plot_crm_skeletons(skeletons, target)
```

# One illustrative CRM trial

```{r one-crm-trial}
example_truth <- scenarios[["MTD at dose 4"]]
example_skeleton <- skeletons[[2L]]

example_trial <- run_crm_trial(
  true_probabilities = example_truth,
  skeleton = example_skeleton,
  target = target,
  n_patients = n_patients,
  cohort_size = cohort_size,
  start_dose = 1L,
  seed = 104L
)

example_trial$selected_dose
example_trial$allocation
```

```{r crm-trial-plots, fig.height=4.5, fig.cap="Allocation path and final posterior toxicity estimates for one CRM trial."}
old_parameters <- par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))
plot_trial_path(example_trial)
title("CRM allocation path")
plot_posterior_toxicity(
  example_trial$fit$posterior_toxicity,
  example_truth,
  target,
  main = "CRM posterior"
)
par(old_parameters)
```

# MCMC posterior calculation

`crm_mcmc()` uses random-walk Metropolis sampling for \(\alpha\). The grid and
MCMC posterior DLT estimates should be similar apart from Monte Carlo error.

```{r mcmc-fit}
mcmc_fit <- crm_mcmc(
  dose = example_trial$history$dose,
  dlt = example_trial$history$dlt,
  skeleton = example_skeleton,
  n_iter = 12000L,
  burn = 2000L,
  thin = 5L,
  proposal_sd = 0.35,
  seed = 902L
)

mcmc_comparison <- data.frame(
  dose = seq_len(n_dose),
  grid_posterior_mean = example_trial$fit$posterior_toxicity,
  mcmc_posterior_mean = mcmc_fit$posterior_toxicity
)
knitr::kable(mcmc_comparison, digits = 3)
mcmc_fit$acceptance_rate
```

```{r mcmc-diagnostics, fig.cap="MCMC trace and retained posterior samples."}
plot_crm_mcmc(mcmc_fit)
```

# BMA-CRM across all skeletons

Yin and Yuan's BMA-CRM regards each skeleton as a model \(M_k\). Its posterior
model probabilities are

\[
\Pr(M_k\mid D)=
\frac{\Pr(M_k)\int L(D\mid\alpha_k,M_k)f(\alpha_k\mid M_k)d\alpha_k}
{\sum_l\Pr(M_l)\int L(D\mid\alpha_l,M_l)f(\alpha_l\mid M_l)d\alpha_l}.
\]

The BMA toxicity estimate is

\[
\widehat\pi_j^{\mathrm{BMA}}
=\sum_k\Pr(M_k\mid D)\widehat\pi_{kj}.
\]

Thus all prespecified skeletons contribute, with more posterior weight assigned
to skeletons that better explain the accumulated DLT observations.

```{r one-bma-trial}
bma_trial <- run_bma_crm_trial(
  true_probabilities = example_truth,
  skeletons = skeletons,
  target = target,
  n_patients = n_patients,
  cohort_size = cohort_size,
  start_dose = 1L,
  seed = 104L
)

bma_weights <- data.frame(
  skeleton = names(skeletons),
  posterior_model_probability = bma_trial$fit$posterior_model_weights
)
knitr::kable(bma_weights, digits = 3)
bma_trial$selected_dose
```

```{r bma-plots, fig.height=4.5, fig.cap="BMA-CRM allocation, evolving model weights, and posterior estimates."}
old_parameters <- par(mfrow = c(1, 3), mar = c(4, 4, 3, 1))
plot_trial_path(bma_trial)
title("BMA-CRM path")
plot_bma_weights(bma_trial)
title("Skeleton weights")
plot_posterior_toxicity(
  bma_trial$fit$posterior_toxicity,
  example_truth,
  target,
  main = "BMA posterior"
)
par(old_parameters)
```

# Simulation study

The study compares each single-skeleton CRM with BMA-CRM under all three true
scenarios. Increase `n_sim` for greater Monte Carlo precision.

```{r simulation-study, cache=TRUE}
n_sim <- 100L

simulation_results <- simulate_crm_designs(
  scenarios = scenarios,
  skeletons = skeletons,
  target = target,
  n_sim = n_sim,
  n_patients = n_patients,
  cohort_size = cohort_size,
  start_dose = 1L,
  seed = 20261004L
)

simulation_summary <- summarize_crm_simulations(
  simulation_results,
  n_dose,
  n_patients
)

operating_table <- simulation_summary$operating
operating_table$probability_correct_selection <- round(
  operating_table$probability_correct_selection, 3
)
operating_table$mean_dlt_rate <- round(operating_table$mean_dlt_rate, 3)
knitr::kable(operating_table)
```

```{r operating-characteristics, fig.height=5.5, fig.cap="Correct MTD selection and observed DLT rate."}
plot_operating_metrics(simulation_summary)
```

```{r selection-probabilities, fig.height=8, fig.cap="Final dose-selection probabilities under each scenario."}
old_parameters <- par(mfrow = c(3, 1), mar = c(4, 4, 3, 1))
for (scenario_name in names(scenarios)) {
  plot_selection_probabilities(simulation_summary, scenario_name)
}
par(old_parameters)
```

```{r allocation-proportions, fig.height=8, fig.cap="Average patient allocation under each scenario."}
old_parameters <- par(mfrow = c(3, 1), mar = c(4, 4, 3, 1))
for (scenario_name in names(scenarios)) {
  plot_allocation_proportions(simulation_summary, scenario_name)
}
par(old_parameters)
```

# Suggested extensions

- Increase `n_sim` to 1,000 and assess Monte Carlo stability.
- Change the prior MTD and indifference-interval half-widths.
- Add a conservative overdose stopping rule.
- Compare cohort sizes of one and three.
- Add a scenario in which every dose is above the target.
- Compare several MCMC proposal standard deviations.

# References

- Lee, S. M. and Cheung, Y. K. (2009). “Model calibration in the continual
  reassessment method.” *Clinical Trials*, 6(3), 227–238.
  [doi:10.1177/1740774509105076](https://doi.org/10.1177/1740774509105076)
- O'Quigley, J., Pepe, M., and Fisher, L. (1990). “Continual reassessment
  method: a practical design for phase I clinical trials in cancer.”
  *Biometrics*, 46(1), 33–48.
- Yin, G. and Yuan, Y. (2009). “Bayesian model averaging continual
  reassessment method in phase I clinical trials.” *Journal of the American
  Statistical Association*, 104(487), 954–968.
  [doi:10.1198/jasa.2009.ap08425](https://doi.org/10.1198/jasa.2009.ap08425)
