---
title: "Continual Reassessment Method: Simulation Practice"
author: ""
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 5,
  fig.align = "center"
)
source("R/crm_functions.R")
```

# Purpose

This coding exercise implements a one-parameter Bayesian continual reassessment
method (CRM). It covers data generation, multiple skeletons, grid and MCMC
posterior computation, sequential dose assignment, BMA-CRM, and operating
characteristics under several true dose-toxicity scenarios.

The code is intentionally transparent and uses base R. It is for learning and
is **not validated software for conducting a clinical trial**.

# Design settings and data-generating scenarios

We use six ordered dose levels, a target DLT probability of 0.25, cohorts of
three, and 24 patients. The true MTD is the dose whose DLT probability is
closest to the target.

```{r design-settings}
target <- 0.25
n_dose <- 6L
n_patients <- 24L
cohort_size <- 3L
prior_mtd <- 3L

scenarios <- list(
  "MTD at dose 2" = c(0.08, 0.23, 0.38, 0.55, 0.68, 0.80),
  "MTD at dose 4" = c(0.04, 0.08, 0.14, 0.24, 0.38, 0.55),
  "MTD at dose 5" = c(0.02, 0.05, 0.09, 0.14, 0.23, 0.36)
)

scenario_table <- data.frame(
  scenario = rep(names(scenarios), each = n_dose),
  dose = rep(seq_len(n_dose), times = length(scenarios)),
  true_dlt_probability = unlist(scenarios, use.names = FALSE)
)
knitr::kable(scenario_table, digits = 2)
```

# CRM working model

For skeleton probabilities \(p_1 < \cdots < p_J\), we use the power model

\[
\Pr(Y=1\mid d_j,\alpha)=p_j^{\exp(\alpha)},
\qquad \alpha\sim N(0,2),
\]

where 2 is the prior variance. After each cohort, the next dose is the level
whose posterior mean DLT probability is closest to the target. Dose movement is
limited to one level per cohort.

Sequential decisions use deterministic grid integration. MCMC is demonstrated
separately, avoiding simulation decisions that depend on Monte Carlo noise.

# Lee-Cheung indifference-interval skeletons

Lee and Cheung (2009) specify an indifference interval
\((\phi-\delta,\phi+\delta)\) around target \(\phi\). Given prior MTD location
\(\nu\), set \(p_\nu=\phi\) and use

\[
p_{j-1}=\exp\left\{
\frac{\log(\phi-\delta)\log(p_j)}{\log(\phi+\delta)}
\right\},
\]

\[
p_{j+1}=\exp\left\{
\frac{\log(\phi+\delta)\log(p_j)}{\log(\phi-\delta)}
\right\}.
\]

A smaller \(\delta\) gives a more closely spaced skeleton. We compare three
prespecified values and assess them through finite-sample simulations.

```{r construct-skeletons}
deltas <- c(0.04, 0.07, 0.10)
skeletons <- lapply(deltas, function(delta) {
  lee_cheung_skeleton(target, delta, n_dose, prior_mtd)
})
names(skeletons) <- paste0("CRM delta=", sprintf("%.02f", deltas))

skeleton_table <- data.frame(
  dose = seq_len(n_dose),
  do.call(cbind, skeletons),
  check.names = FALSE
)
knitr::kable(skeleton_table, digits = 3)
```

```{r skeleton-plot, fig.cap="Three Lee-Cheung skeletons."}
plot_crm_skeletons(skeletons, target)
```

# One illustrative CRM trial

```{r one-crm-trial}
example_truth <- scenarios[["MTD at dose 4"]]
example_skeleton <- skeletons[[2L]]

example_trial <- run_crm_trial(
  true_probabilities = example_truth,
  skeleton = example_skeleton,
  target = target,
  n_patients = n_patients,
  cohort_size = cohort_size,
  start_dose = 1L,
  seed = 104L
)

example_trial$selected_dose
example_trial$allocation
```

```{r crm-trial-plots, fig.height=4.5, fig.cap="Allocation path and final posterior toxicity estimates for one CRM trial."}
old_parameters <- par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))
plot_trial_path(example_trial)
title("CRM allocation path")
plot_posterior_toxicity(
  example_trial$fit$posterior_toxicity,
  example_truth,
  target,
  main = "CRM posterior"
)
par(old_parameters)
```

## Clinical interpretation of truth and posterior mean

The **truth** line: For example, a true probability of 0.24 at dose 4 means that
each simulated patient assigned to dose 4 has a 24% chance of a DLT. 

The **posterior mean** line is the CRM's current estimate after combining the
skeleton, the prior for \(\alpha\), and the DLT observations. At dose \(j\), it
is

\[
\widehat\pi_j=E\left[p_j^{\exp(\alpha)}\mid D\right]
=\int p_j^{\exp(\alpha)}f(\alpha\mid D)d\alpha.
\]

Clinically, if the posterior mean is below the truth, the
model is underestimating toxicity.Either discrepancy can change which dose appears closest
to the target. A real protocol should therefore also use prespecified safety
rules and posterior uncertainty, not just the posterior mean.

# MCMC posterior calculation

`crm_mcmc()` uses random-walk Metropolis sampling for \(\alpha\). The grid and
MCMC posterior DLT estimates should be similar apart from Monte Carlo error.

## Grid integration versus MCMC

Both methods target the same posterior distribution; they differ only in how
the integral is approximated.

- **Grid integration** evaluates the posterior kernel at 801 fixed values of
  \(\alpha\) between -4 and 4, normalizes their weights, and calculates
  \(\widehat\pi_j\) as a weighted sum. For this one-parameter model it is fast and stable.

- **MCMC** draws samples  from the posterior distribution\(f(\alpha\mid D)\). It estimates the same quantity with
  \(S^{-1}\sum_{s=1}^S p_j^{\exp(\alpha_s)}\). 

Grid integration is therefore used for the repeated interim decisions in this
exercise. 

```{r mcmc-fit}
mcmc_fit <- crm_mcmc(
  dose = example_trial$history$dose,
  dlt = example_trial$history$dlt,
  skeleton = example_skeleton,
  n_iter = 12000L,
  burn = 2000L,
  thin = 5L,
  proposal_sd = 0.35,
  seed = 902L
)

mcmc_comparison <- data.frame(
  dose = seq_len(n_dose),
  grid_posterior_mean = example_trial$fit$posterior_toxicity,
  mcmc_posterior_mean = mcmc_fit$posterior_toxicity
)
knitr::kable(mcmc_comparison, digits = 3)
mcmc_fit$acceptance_rate
```

```{r mcmc-diagnostics, fig.cap="MCMC trace and retained posterior samples."}
plot_crm_mcmc(mcmc_fit)
```

# BMA-CRM across all skeletons

Yin and Yuan's BMA-CRM regards each skeleton as a model \(M_k\). Its posterior
model probabilities are

\[
\Pr(M_k\mid D)=
\frac{\Pr(M_k)\int L(D\mid\alpha_k,M_k)f(\alpha_k\mid M_k)d\alpha_k}
{\sum_l\Pr(M_l)\int L(D\mid\alpha_l,M_l)f(\alpha_l\mid M_l)d\alpha_l}.
\]

The BMA toxicity estimate is

\[
\widehat\pi_j^{\mathrm{BMA}}
=\sum_k\Pr(M_k\mid D)\widehat\pi_{kj}.
\]

Thus all prespecified skeletons contribute, with more posterior weight assigned
to skeletons that better explain the accumulated DLT observations.

```{r one-bma-trial}
bma_trial <- run_bma_crm_trial(
  true_probabilities = example_truth,
  skeletons = skeletons,
  target = target,
  n_patients = n_patients,
  cohort_size = cohort_size,
  start_dose = 1L,
  seed = 104L
)

bma_weights <- data.frame(
  skeleton = names(skeletons),
  posterior_model_probability = bma_trial$fit$posterior_model_weights
)
knitr::kable(bma_weights, digits = 3)
bma_trial$selected_dose
```

```{r bma-plots, fig.height=4.5, fig.cap="BMA-CRM allocation, evolving model weights, and posterior estimates."}
old_parameters <- par(mfrow = c(1, 3), mar = c(4, 4, 3, 1))
plot_trial_path(bma_trial)
title("BMA-CRM path")
plot_bma_weights(bma_trial)
title("Skeleton weights")
plot_posterior_toxicity(
  bma_trial$fit$posterior_toxicity,
  example_truth,
  target,
  main = "BMA posterior"
)
par(old_parameters)
```

# Simulation study

The study compares each single-skeleton CRM with BMA-CRM under all three true
scenarios. Increase `n_sim` for greater Monte Carlo precision.

## How the simulation probabilities are calculated

For scenario \(s\), let \(\pi_{sj}\) be the true DLT probability at dose \(j\).
The simulation defines the true MTD as

\[
j_s^*=\arg\min_j |\pi_{sj}-\phi|.
\]

Within simulated trial \(r\), binary outcomes are independently generated as

\[
Y_{ir}\sim\operatorname{Bernoulli}(\pi_{s,d_{ir}}),
\]

where \(d_{ir}\) is the dose assigned by the evolving CRM. At the end of the
trial, the selected dose is

\[
\widehat j_{sr}=\arg\min_j|\widehat\pi_{srj}-\phi|,
\]

using either the single-skeleton posterior mean or the BMA posterior mean. With
\(R\) simulated trials, the reported metrics are:

\[
\operatorname{PCS}_s=\frac{1}{R}\sum_{r=1}^R
I(\widehat j_{sr}=j_s^*),
\]

\[
\Pr(\text{select dose }j)=\frac{1}{R}\sum_{r=1}^R
I(\widehat j_{sr}=j),
\]

\[
\text{mean DLT rate}=\frac{1}{Rn}\sum_{r=1}^R\sum_{i=1}^nY_{ir},
\qquad
\text{allocation proportion}_j=
\frac{1}{Rn}\sum_{r=1}^R\sum_{i=1}^nI(d_{ir}=j).
\]

These are Monte Carlo estimates. With only 100 simulations they will vary
somewhat if the seed changes; a confirmatory design study would normally use
many more replicates and report Monte Carlo standard errors.

```{r simulation-study, cache=TRUE}
n_sim <- 1000L

simulation_results <- simulate_crm_designs(
  scenarios = scenarios,
  skeletons = skeletons,
  target = target,
  n_sim = n_sim,
  n_patients = n_patients,
  cohort_size = cohort_size,
  start_dose = 1L,
  seed = 20261004L
)

simulation_summary <- summarize_crm_simulations(
  simulation_results,
  n_dose,
  n_patients
)

operating_table <- simulation_summary$operating
operating_table$probability_correct_selection <- round(
  operating_table$probability_correct_selection, 3
)
operating_table$mean_dlt_rate <- round(operating_table$mean_dlt_rate, 3)
knitr::kable(operating_table)
```

```{r operating-characteristics, fig.height=5.5, fig.cap="Correct MTD selection and observed DLT rate."}
plot_operating_metrics(simulation_summary)
```

```{r selection-probabilities, fig.height=8, fig.cap="Final dose-selection probabilities under each scenario."}
old_parameters <- par(mfrow = c(3, 1), mar = c(4, 4, 3, 1))
for (scenario_name in names(scenarios)) {
  plot_selection_probabilities(simulation_summary, scenario_name)
}
par(old_parameters)
```

## Three starting-dose allocation plans: 3 x 3 x 3 comparison

CRM does not fix the final number of patients at each dose. After an initial
starting-dose group, allocation adapts to the observed DLTs. A practical design
choice is how much safety information to collect at the starting dose before
the first model-based reassessment. We compare:

1. **3-patient run-in:** the current design; update after the first cohort;
2. **6-patient run-in:** observe six patients at dose 1 before adapting; and
3. **9-patient run-in:** a deliberately cautious plan with nine patients at
   dose 1 before adapting.

All plans retain 24 total patients and use cohorts of three after the run-in.
Larger run-ins should put more patients at early doses and may reduce early
toxicity, but they leave fewer patients for learning near a higher true MTD.
They illustrate a clinically meaningful safety-efficiency tradeoff rather than
three fixed allocations.

Crossing 3 run-in plans, 3 true scenarios, and 3 skeletons produces 27 settings.
BMA-CRM is left out of this factorial display because it combines the three
skeletons rather than representing a fourth skeleton.

```{r run-in-factorial-simulation, cache=TRUE}
run_in_sizes <- c(3L, 6L, 9L)

factorial_results <- simulate_crm_run_in_plans(
  scenarios = scenarios,
  skeletons = skeletons,
  target = target,
  run_in_sizes = run_in_sizes,
  n_sim = n_sim,
  n_patients = n_patients,
  cohort_size = cohort_size,
  start_dose = 1L,
  seed = 20261005L
)

factorial_summary <- summarize_crm_run_in_plans(
  factorial_results,
  n_patients
)
factorial_table <- factorial_summary
numeric_columns <- c(
  "probability_correct_selection",
  "mean_dlt_rate",
  "mean_low_dose_allocation"
)
factorial_table[numeric_columns] <- lapply(
  factorial_table[numeric_columns], round, digits = 3
)
factorial_table <- factorial_table[order(
  factorial_table$design,
  factorial_table$scenario,
  factorial_table$run_in_size
), ]
knitr::kable(factorial_table, row.names = FALSE)
```

Each heatmap panel is one skeleton. Rows are the three true scenarios, columns
are the three run-in plans, and each printed cell is the estimated metric.

```{r run-in-pcs-heatmap, fig.width=12, fig.height=5.5, fig.cap="Probability of correct MTD selection across the 27 settings."}
plot_crm_factorial_heatmaps(
  factorial_summary,
  metric = "probability_correct_selection",
  title = "Probability of correct MTD selection"
)
```

```{r run-in-allocation-heatmap, fig.width=12, fig.height=5.5, fig.cap="Mean proportion assigned to dose levels 1 or 2 across the 27 settings."}
plot_crm_factorial_heatmaps(
  factorial_summary,
  metric = "mean_low_dose_allocation",
  title = "Mean allocation to dose levels 1 or 2"
)
```

```{r run-in-dlt-heatmap, fig.width=12, fig.height=5.5, fig.cap="Mean observed DLT rate across the 27 settings."}
plot_crm_factorial_heatmaps(
  factorial_summary,
  metric = "mean_dlt_rate",
  title = "Mean observed DLT rate",
  zlim = c(0, max(factorial_summary$mean_dlt_rate))
)
```

```{r allocation-proportions, fig.height=8, fig.cap="Average patient allocation under each scenario."}
old_parameters <- par(mfrow = c(3, 1), mar = c(4, 4, 3, 1))
for (scenario_name in names(scenarios)) {
  plot_allocation_proportions(simulation_summary, scenario_name)
}
par(old_parameters)
```

# Suggested extensions

- Increase `n_sim` to 1,000 and assess Monte Carlo stability.
- Change the prior MTD and indifference-interval half-widths.
- Add a conservative overdose stopping rule.
- Compare cohort sizes of one and three.
- Add a scenario in which every dose is above the target.
- Compare several MCMC proposal standard deviations.
